Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
Updates

\lstset{
      language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
      tabsize=2,
      escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
      breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
      keywords=[2]{},
      keywordstyle=[2]\color{byzantine}\bfseries,
      keywords=[3]{eff,effect,perform,continue,discontinue,continuation,
                   clone_continuation},
      keywordstyle=[3]\color{caribbeangreen}\bfseries,
      keywords=[4]{call,return,external},
      keywordstyle=[4]\color{Bittersweet}\bfseries,
      classoffset=1,
      upquote=true,
      keywordstyle=\color{byzantine}\bfseries,
      classoffset=0,
      mathescape=true,
      numberstyle=\tiny\color{gray},
      numbersep=5pt
}

@@

\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}
\setlength{\tabcolsep}{3pt}

@@

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
  continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicore OCaml'' to
distinguish it from stock OCaml.

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:refine}, we refine the design further focussing on
the issues that come up when integrating effect handlers into a mainstream
systems language. \S\ref{sec:semantics} presents the formal operational
semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
compiler and the runtime system support for implementing effect handlers.
\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
\S\ref{sec:conc} present the related work and the conclusions.

@@

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.46]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
  \mycaption{Program stack on stock OCaml.}
\end{figure*}

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. In this section, we provide an overview of the
program stack and related mechanisms in stock OCaml.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the OCaml runtime
makes use of C callee-saved registers: |r15| holds the \emph{allocation pointer}
into the minor heap used for bump pointer allocation, and |r14| holds a
reference to the |Caml_state| table of global variables used by the runtime.
This makes external calls extremely fast in OCaml. If the external function does
not allocate in the OCaml heap, then it can be called directly and no
bookkeeping is necessary. For external functions which allocate in the OCaml
heap, the cached allocation pointer is saved to |Caml_state| before the external
call and it is restored on return. Similarly, callbacks into OCaml are also
cheap: these involve loading the arguments in the right registers and calling
the OCaml function. OCaml callbacks are relatively common as the garbage
collector (GC) finalisation functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Installing an exception handler simply pushes the program counter (|pc|) of the
handler and the current exception pointer (|exn_ptr| -- a field in |Caml_state|)
to the stack. At this point, the current exception pointer is updated to be the
current stack pointer (|rsp|). This creates a linked-list of exception handler
frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
to the |pc| of the handler.

@@

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.
Note that the DWARF unwind information is complementary to the information used
by OCaml to walk the stack for GC.

In summary, the design of effect handlers in OCaml must accommodate these
desirable properties of OCaml with respect to the program stack.

@@

\section{Semantics}
\label{sec:semantics}

In this section, we formalise the effect handler design in Multicore OCaml.

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

@@

@@

	\vspace{-3mm}
  \caption{Syntax of expressions and configurations}

@@

	\vspace{-3mm}

@@

	\mycaption{Top-level reductions}

@@

  \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}

@@

  \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}

@@

	\vspace{-3mm}

@@

We present the operational semantics for a core language of effect handlers.
The operational semantics intends to faithfully capture the semantics of
Multicore OCaml effect handlers, and to inform an implementation.

\subsubsection{Syntax}

Our expressions (Figure~\ref{sem:syntax}) consists of integer constants ($n$),
variables ($x$), abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic
expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
consists of a return case ($\caseval{x}{e}$), zero or more exception cases
($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
continuation $k$ and body $e$.

@@

\subsubsection{Top-level reductions}

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.

\subsubsection{C reductions}

We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
reduction step $\rightsquigarrow$. The administrative reductions are common to
both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
a function frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. We shall return to \textsc{Resume1}
and \textsc{Resume2} later.

@@

\subsubsection{OCaml reductions}

\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcll}
      \textsc{AdminO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
                        & \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
                        & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
      \textsc{RetToC}   & (v, \_, [([],(h,\emptyset))], \cstack) & \ostep
                        & \configa{v}{\env}{\cstack}
                        & \text{if } h = \{\caseval{x}{x}\} \\
      \textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstack}}
                        & \text{if } \{\caseval{x}{e}\} \in h \\
      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
      \textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
                        & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
      \textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
    \end{array}
  \end{smathpar}
  \vspace{-1mm}
  \mycaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\end{figure*}

In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
fiber in the current stack. \textsc{AdminO} performs administrative reductions.
\textsc{CallO} evaluates an OCaml function on the current stack.
\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
on an empty C stack. \textsc{RetToC} returns a value to the enclosing C stack.
In this case, we have exactly one fiber on the stack, and this was created in
the rule \textsc{Callback}, whose handler has identity return case alone and
the environment is empty. \textsc{RetFib} returns value from a fiber to the
previous one, evaluating the body of the return case.

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
\vspace{-3mm}
\end{figure*}

@@

Multicore OCaml is not the first language to support stack inspection in the
presence of non-local control operators. Chez Scheme supports continuation
marks~\cite{Flatt20} which permit stack inspection to implement dynamic
binding, exceptions, profilers, debuggers, etc, in the presence of first-class
continuations. As the authors note, continuation marks can be implemented using
effect handlers, but direct support for continuation marks leads to better
performance. In this work, we focus on retaining the support for stack
inspection through DWARF unwind tables in the presence of effect handlers.

@@

>>>>

\lstset{
      language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
      tabsize=2,
      escapechar={<@}{@>},
      basewidth={0.5em,0.45em},
      aboveskip={3pt},
      belowskip={3pt},
      breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
      keywords=[2]{},
      keywordstyle=[2]\color{byzantine}\bfseries,
      keywords=[3]{eff,effect,perform,continue,discontinue,continuation,
                   clone_continuation},
      keywordstyle=[3]\color{caribbeangreen}\bfseries,
      keywords=[4]{call,return,external},
      keywordstyle=[4]\color{Bittersweet}\bfseries,
      classoffset=1,
      upquote=true,
      keywordstyle=\color{byzantine}\bfseries,
      classoffset=0,
      mathescape=true,
      numberstyle=\tiny\color{gray},
      numbersep=5pt
}

@@

\setlength{\tabcolsep}{3pt}

@@

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string (perform (Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
		register_async_input_line (chan, k);
		continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation \textbf{Multicore OCaml} to
distinguish it from stock OCaml.

The rest of the paper is organized as follows. The next section describes the
details of stock OCaml program stack. In \S\ref{sec:refine}, we refine the
design further focussing on retrofitting effect handlers into a mainstream
\emph{systems} language. \S\ref{sec:semantics} presents the compilation scheme
and operational semantics semantics for Multicore OCaml effect handlers.
\S\ref{sec:impl} discusses the compiler and the runtime system support for
implementing effect handlers. \S\ref{sec:eval} presents extensive performance
evaluation of effect handlers against the goals set in \S\ref{sec:req}.
\S\ref{sec:related} and \S\ref{sec:conc} present the related work and the
conclusions.

@@

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.44]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
\vspace{-2mm}
\caption{Program stack on stock OCaml.}
\vspace{-2mm}
\end{figure*}

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack and preserving its desirable properties. In this
section, we provide an overview of the program stack and related mechanisms in
stock OCaml.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
makes use of two C callee-saved registers for supporting OCaml execution. The
register |r15| holds the \emph{allocation pointer} into the minor heap used for
bump pointer allocation, and |r14| holds a reference to the |Caml_state| table
of global variables used by the runtime. This makes external calls extremely
fast in OCaml. If the external function does not allocate in the OCaml heap,
then it can be called directly and no bookkeeping is necessary. For external
functions which allocate in the OCaml heap, the cached allocation pointer is
saved to |Caml_state| before the external call and it is restored on return.
Similarly, callbacks into OCaml are also cheap: these involve loading the
arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Installing an exception handler simply pushes the program counter (|pc|) of the
handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|) to the stack. After this, the current exception pointer is
updated to be the current stack pointer (|rsp|). This creates a linked-list of
exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
and jumps to the |pc| of the handler.

@@

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.
Note that the DWARF unwind information is complementary to the information used
by OCaml to walk the stack for GC.

@@

\vspace{-2mm}
\section{Semantics}
\label{sec:semantics}

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

@@

\begin{minipage}{\linewidth}

@@

  \subcaption{Syntax of expressions and configurations}

@@

%
\begin{minipage}{\linewidth}

@@

	\subcaption{Top-level reductions}

@@

  \subcaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}

@@

  \subcaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}

@@

%
\begin{minipage}{\linewidth}
      \textsc{AdminO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
                        & \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
      \textsc{RetToC}   & (v, \_, [([],(h,\emptyset))], \cstack) & \ostep
                        & \configa{v}{\env}{\cstack}
                        & \text{if } h = \{\caseval{x}{x}\} \\
      \textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstack}}
                        & \text{if } \{\caseval{x}{e}\} \in h \\
      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
      \textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
                        & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
      \textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\caption{Operational semantics of Multicore OCaml effect handlers.}

@@

We present an operational semantics for a core language of effect handlers that
faithfully captures the semantics of the Multicore OCaml implementation.

\textbf{\textit{Syntax.}} Our expressions (Figure~\ref{sem:syntax}) consists of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

\textbf{\textit{Top-level reductions.}} The initial configuration for an expression
$e$ is $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment
and the stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel})
can be performed by either by taking a C step $\cstep$ or an OCaml step
$\ostep$.

\textbf{\textit{C reductions.}} We can take a C step (Figure~\ref{sem:cstep}) by
taking an administrative reduction step $\rightsquigarrow$. The administrative
reductions are common to both C and OCaml. The rules \textsc{Var},
\textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2} and
\textsc{App3} are standard. \textsc{Arith3} performs the arithmetic operation
on the integers ($\llbracket n_1 \odot n_2 \rrbracket$). \textsc{Raise} pushes
an function frame with exception value to indicate that an exception is being
raised. Similarly, \textsc{Perform} pushes a function frame with an effect
value with an empty continuation $[[],(\{\caseval{x}{x}\},\emptyset)]$ with no
captured frames and an empty handler with an identity return case alone. We
shall return to \textsc{Resume1} and \textsc{Resume2} later.

@@

\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}), reductions
always occur on the top-most fiber in the current stack. \textsc{AdminO}
performs administrative reductions. \textsc{CallO} evaluates an OCaml function
on the current stack. \textsc{ExtCall} captures the behaviour of external
calls, which are evaluated on an empty C stack. \textsc{RetToC} returns a value
to the enclosing C stack. In this case, we have exactly one fiber on the stack,
and this was created in the rule \textsc{Callback}, whose handler has identity
return case alone and the environment is empty. \textsc{RetFib} returns value
from a fiber to the previous one, evaluating the body of the return case.

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
	with | v -> v | effect F kf -> ()
with | v -> v
		 | effect E ke -> (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
\vspace{-3mm}
\end{figure*}

@@

Multicore OCaml is not the first language to support stack inspection in the
presence of non-local control operators. Chez Scheme supports continuation
marks~\cite{Flatt20} which permit stack inspection as a language feature. This
enables implementation of dynamic binding, exceptions, profilers, debuggers,
etc, in the presence of first-class continuations. As the authors note,
continuation marks can be implemented using effect handlers, but direct support
for continuation marks leads to better performance. In this work, we focus on
retaining the support for stack inspection through DWARF unwind tables in the
presence of effect handlers.

@@

\subsection*{Experimental support for multi-shot continuations}
